Reducts of relation algebras: finite representability and recursively enumerable axiomatisability

### 2021-11-08 frp-paper/Paper.tex
cylindric notes, wip

\author{Daniel Rogozin \\ Institute for Information Transmission Problems, Russian Academy of Sciences \\ daniel.rogozin@serokell.io}
\date{}
\title{The finite representation property for some reducts of relation algebras}
\providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}

@@

We also show that the class of representable join semilattice-ordered semigroups has a recursively
enumerable axiomatisation and the finite representation property.
\end{abstract}

@@

Relation algebras are the kind of Boolean algebras with operators that provide algebraisation
of binary relations \cite{jonsson1951boolean}. The class of all relation algebras,
denoted as ${\bf RA}$, consists of algebras the signature of which is $\{ 0, 1, +, -, ;, {}^{\smile}, {\bf 1}' \}$ obeying certain axioms.
The class of representable relation algebras, ${\bf RRA}$, consists of algebras isomorphic to set relation algebras. ${\bf RRA}$ is a subclass of ${\bf RA}$, but the converse inclusion does not hold.
That is, there exist relation algebras with no representation as set relation algebras
\cite{lyndon1950representation}. Moreover, the class ${\bf RRA}$ is not finitely axiomatisable \cite{monk1964representable}. The problem of determining whether a given finite relation algebra
is representable is undecidable, see \cite{hirsch2001representability}.

In other words, relation algebras are quite badly behaved. The study of such reducts is mostly motivated by such ``bad behaviour'' in order to avoid these restrictions and determine the possible reasons for them.

There are several results on reducts of relation algebras having no finite axiomatisation such as ordered monoids \cite{hirsch2005class}, distributive residuated lattices \cite{andreka1994lambek}, join semilattice-ordered semigroups \cite{andreka2011axiomatizability}, algebras whose signature contains composition, meet, and converse \cite{hodkinson2000axiomatizability}, etc.

@@

In the last section, we study the class of representable join semilattice-ordered semigroups, denoted as ${\bf R}(;,+)$. It is
already known that this class is not finitely axiomatisable \cite{andreka2011axiomatizability}.
First, we show that ${\bf R}(;,+)$ has a recursively enumerable axiomatisation. For that,
we define networks and representability games for them. This class is axiomatised with the axioms of join semilattice-ordered semigroups plus the countable set of universal formulas claiming that
$\exists$ has a winning strategy on every finitary step.

After that, we show that ${\bf R}(;,+)$ has the finite representation property. Moreover, we extract the explicit recursively enumerable axiomatisation of ${\bf R}(;,+)$. As a corollary, the representability problem for finite join semilattice-ordered semigroup is decidable.

@@

Let us the basic definitions related to relation algebras \cite{hirsch2002relation}.
\begin{definition} A relation algebra is an algebra $\mathcal{R} = \langle R, 0, 1, +, -, ;, {}^{\smile}, {\bf 1 }\rangle$ such that $\langle R, 0, 1, +, - \rangle$ is a Boolean algebra and the following equations hold, for each $a, b, c \in R$:
    \begin{enumerate}
      \item $a ; (b ; c) = (a ; b) ; c$,
      \item $(a + b) ; c = (a ; c) + (b ; c)$,
      \item $a ; {\bf 1}' = a$,
      \item $a^{\smile \smile} = a$,
      \item $(a + b)^{\smile} = a^{\smile} + b^{\smile}$,
      \item $(a ; b)^{\smile} = b^{\smile} ; a^{\smile}$,
      \item $a^{\smile} ; (- (a ; b)) \leq - b$.
    \end{enumerate}
Note that $a \leq b$ iff $a + b = b$ iff $a \cdot b = a$, where $a \cdot b = - (- a + - b)$. ${\bf RA}$ is the class of all relation algebras.
\end{definition}

@@

@@

\begin{center}
$(\mathcal{P}(A))_m = \{ X \in \mathcal{P}(S) \: | \: m X = X\}$ )
\end{center}
forms a complete lattice with $\Sigma_{m} \mathcal{X} = m ( \bigcup \mathcal{X})$ and $\Pi_{m} = \bigcap \mathcal{X}$, see \cite[Theorem 7.3]{davey2002introduction}.

@@

\section{Join semilattice-ordered semigroups: the explicit axiomatisation and the finite representation property}

We  note  that  a  similar  construction  does  not  work  for  finite  representableupper semilattice-ordered semigroups. From the one hand, the notions of afinite upper semilattice-ordered semigroup and finite quantale are quite close to each other. From the other hand, there are examples of non-representable upper semilattice-ordered semigroups whose any non-principal ultraproduct is representable, see \cite[Theorem 3.1]{andreka2011axiomatizability}. That makes ${\bf R}(;,+)$ not finitely axiomatisable. Thus, one needs to show the decidability of membership in a different way. We have the following fact:

In this section, we show that ${\bf R}(+, ;)$ has a recursively enumerable universal axiomatisation. For that, we characterise representability using Ehrenfeucht-Fra\"{i}ss\'{e} games on networks. After that, we show that every finite $\mathcal{A} \in {\bf R}(+, ;)$ is representable by factorising its network through an equivalence relation having a finite index. That is, an equivalence relation whose quotient is finite.

@@

\subsection{The finite representation property for ${\bf R}(+, ;)$}

Let us show finite representability for ${\bf R}(+, ;)$. Suppose that a finite $\mathcal{A}$ is representable, so $\exists$ has a winning strategy on $\mathcal{G}^{\omega}(\mathcal{A})$. Let us consider $\mathcal{N} = \langle V, E, l \rangle$, the network constructed during the $\mathcal{G}^{\omega}(\mathcal{A})$ game.

\begin{definition}
  Let $\mathcal{A}$ be a join semilattice-ordered semigroup and let $\mathcal{N}_1$ and $\mathcal{N}_2$ be $\mathcal{A}$-networks. A strong homomorphism from $\mathcal{N}_1$ onto $\mathcal{N}_2$ is a map $f : \mathcal{N}_1 \twoheadrightarrow \mathcal{N}_2$ such that:
  \begin{enumerate}
    \item $(x, y) \in E_2$ implies $(f(x), f(y)) \in E_1$,
    \item If $(x', y') \in E_2$, then there exists $x, y \in E_1$ such that $f(x) = x'$ and $f(y) = y'$ and $(x, y) \in E_1$,
    \item $l_{\mathcal{N}_1}(x, y) = l_{\mathcal{N}_2}(f(x),f(y))$
  \end{enumerate}
  By $\mathcal{N}_1 \twoheadrightarrow \mathcal{N}_2$ we mean that there exists a strong homomorphism of $\mathcal{N}_1$ onto $\mathcal{N}_2$.
\end{definition}

\begin{lemma}
  Let $\mathcal{A}$ be a representable join semilattice-ordered semigroup and $\mathcal{N} = \langle V, E, l \rangle$ and $\mathcal{N}' = \langle V', E', l' \rangle$ $\mathcal{A}$-networks.

  Suppose $\mathcal{N} \twoheadrightarrow \mathcal{N}'$. If $\mathcal{N}^{*}$ is a representation of $\mathcal{A}$, so is $ $
\end{lemma}

\begin{lemma}
\end{lemma}

@@

The author is sincerely grateful to Robin Hirsch, Ian Hodkinson, and his supervisor Ilya Shapirovsky for valuable comments. The previous version of this paper contained an error. Many thanks to Ja\v{s} \v{S}emrl for noticing that.

>>>>

\author{Daniel Rogozin \\ Institute for Information Transmission Problems, Russian Academy of Sciences \\ daniel.rogozin@serokell.io}
\date{}
\title{Reducts of relation algebras: finite representability and recursively enumerable axiomatisability}
\providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}

@@

We also show that the class of representable join semilattice-ordered semigroups has a recursively
enumerable axiomatisation using back-and-forth games.
\end{abstract}

@@

Relation algebras are a kind of Boolean algebras with operators that provide algebraisation
of binary relations \cite{jonsson1951boolean}. The class of all relation algebras,
denoted as ${\bf RA}$, consists of algebras of the signature of which is $\{ 0, 1, +, -, ;, {}^{\smile}, {\bf 1}' \}$, and all those algebras obey certain axioms.
The class of representable relation algebras, ${\bf RRA}$, consists of algebras isomorphic to set relation algebras. ${\bf RRA}$ is a subclass of ${\bf RA}$, but the converse inclusion does not hold.
That is, there exist relation algebras with no representation as set relation algebras
\cite{lyndon1950representation}. Moreover, the class ${\bf RRA}$ is not a finitely axiomatisable variety \cite{monk1964representable} with neither Sahlqvist \cite{venema1997atom} nor axiomatisation \cite{hodkinson2005canonical}. The problem of determining whether a given finite relation algebra
is representable is undecidable, see \cite{hirsch2001representability}.

In other words, relation algebras are quite badly behaved. For this reason, we are interested in reduct since they might behave better.

There are several results on reducts of relation algebras that have no finite axiomatisation such as ordered monoids \cite{hirsch2005class}, distributive residuated lattices \cite{andreka1994lambek}, join semilattice-ordered semigroups \cite{andreka2011axiomatizability}, algebras whose signature contains composition, meet, and converse \cite{hodkinson2000axiomatizability}, etc.

@@

In the last section, we study the class of representable join semilattice-ordered semigroups, denoted as ${\bf R}(;,+)$. It is
already known that this class is not finitely axiomatisable \cite{andreka2011axiomatizability}.
We show that ${\bf R}(;,+)$ has a recursively enumerable axiomatisation. For that,
we define networks and representability games for them. This class is axiomatised with the axioms of join semilattice-ordered semigroups plus the countable set of universal formulas claiming that
$\exists$ has a winning strategy on every finitary step. The question of finite representability for this class remains open \cite{vsemrl2021domain}.

@@

Let us introduce the basic definitions related to relation algebras. See \cite[Section 3]{hirsch2002relation} to have more details.
\begin{definition} A relation algebra is an algebra $\mathcal{R} = \langle R, 0, 1, +, -, ;, {}^{\smile}, {\bf 1 }\rangle$ such that $\langle R, 0, 1, +, - \rangle$ is a Boolean algebra and the following equations hold, for each $a, b, c \in R$:
    \begin{enumerate}
      \item $a ; (b ; c) = (a ; b) ; c$,
      \item $(a + b) ; c = (a ; c) + (b ; c)$,
      \item $a ; {\bf 1}' = a$,
      \item $a^{\smile \smile} = a$,
      \item $(a + b)^{\smile} = a^{\smile} + b^{\smile}$,
      \item $(a ; b)^{\smile} = b^{\smile} ; a^{\smile}$,
      \item $a^{\smile} ; (- (a ; b)) \leq - b$.
    \end{enumerate}
Note that $a \leq b$ iff $a + b = b$ iff $a \cdot b = a$, where $a \cdot b = - (- a + - b)$. ${\bf RA}$ is the class of all relation algebras.
\end{definition}

@@

The finite representation property and recursively enumerable axiomatisability imply decidability for finite structures in the following sense. One may think of that as a counterpart of Harrop's theorem from universal algebra \cite{harrop1958existence}.

@@

\begin{center}
$(\mathcal{P}(A))_m = \{ X \in \mathcal{P}(S) \: | \: m X = X\}$
\end{center}
forms a complete lattice with $\Sigma_{m} \mathcal{X} = m ( \bigcup \mathcal{X})$ and $\Pi_{m} = \bigcap \mathcal{X}$, see \cite[Theorem 7.3]{davey2002introduction}.

@@

\section{Join semilattice-ordered semigroups: the explicit axiomatisation}

We  note  that  a  similar  construction  does  not  work  for  finite  representableupper semilattice-ordered semigroups. From the one hand, the notions of afinite upper semilattice-ordered semigroup and finite quantale are quite close to each other. From the other hand, there are examples of non-representable upper semilattice-ordered semigroups whose any non-principal ultraproduct is representable, see \cite[Theorem 3.1]{andreka2011axiomatizability}. That makes ${\bf R}(;,+)$ not finitely axiomatisable. Although, this class has a universal recursively enumerable axiomatisation.

In this section, we show that ${\bf R}(+, ;)$ has a recursively enumerable universal axiomatisation. For that, we characterise representability using Ehrenfeucht-Fra\"{i}ss\'{e} games on networks.

@@

@@

The author is sincerely grateful to Robin Hirsch, Ja\v{s} \v{S}emrl, and his supervisor Ilya Shapirovsky for valuable comments.
